Influence of Time Integration Methodology on Double Detonation Type Ia Supernova Nucleosynthesis

### 2023-06-25 subch_tols/paper.tex
more work

\citet{castro_simple_sdc} looked at an extreme version of the
double detonation scenario---a very large perturbation was applied
to drive a detonation directly into the underlying
carbon-oxygen white dwarf.  We showed in that paper that the Strang
split integration method had difficulty with the integration unless
we used tighter tolerances

Summary of the SDC subch results form the original paper

@@

We use a 2d axisymmetric domain with a size of $5.12\times 10^9~\mathrm{cm}$ by $1.024\times 10^{10}~\mathrm{cm}$.  This is
much larger than the initial size of the star, giving it plenty of
room to expand.  The coarse grid is $640\times 1280$ zones and we use
2 levels of refinement, giving a maximum resolution of
$20~\mathrm{km}$.  The refinement strategy is picked to refine on the
star and any regions where the temperature is greater than
$10^8~\mathrm{K}$.  \castro\ uses subcycling, so finer grids are
advanced at a smaller timestep than the coarse grids.  Self-gravity is
done using a full Poisson solve using geometric multigrid, with
Dirichlet boundary conditions on the domain boundary computed via a
multipole expansion with a maximum order of 6.  For the low density
regions outside of the star, we use a sponge term on the momentum
equation to prevent the very low density material that is not in
hydrostatic equilibrium from raining down on the star.

Our default CFL number is 0.2 for these simulations, and the CFL constraint
is the only timestep constraint used.  For larger CFL numbers, the ODE integration
sometimes encounters problems \MarginPar{more}.

All simulations use the same network, the {\tt subch\_simple} as
described in \cite{zhi2023}.  This includes 22 nuclei and 94 rates
from ReacLib \citep{reaclib} and is produced with \pynucastro\
\citep{pynucastro2}.  Importantly, we include
$\isotm{N}{14}(\alpha,\gamma)\isotm{F}{18}(\alpha,p)\isotm{Ne}{21}$
which creates protons to allow for
$\isotm{C}{12}(p,\gamma)\isotm{N}{13}(\alpha,p)\isotm{O}{16}$.  This
sequence can be faster than
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$, as pointed out by
\citet{shenbildsten} and is important for getting the detonation speed
correctly.  To reduce the size of the network, some
$(\alpha,p)(p,\gamma)$ and $(\alpha,\gamma)$ rates are combined into
an effective $(\alpha,\gamma)$ rate.  Screening is provided following
the procedure in \citet{wallace:1982}, combining the screening
functions of \citet{graboske:1973,alastuey:1978,itoh:1979}.

The reaction system is integrated using a version of the VODE ODE integrator
\citep{vode} ported to C++.  Our modifications to this integrator are described
in \citet{castro_simple_sdc}.  We use an analytic approximation to the Jacobian
and found that we get best results when we disable the caching of the Jacobian
in the integrator.

@@

We will run a suite of simulations all using the same initial model.
The initial model is constructed following the methodology in
\citet{subchandra}.  We use an isothermal core of C/O of $1.1~M_\odot$
with $T = 10^7~\mathrm{K}$ and a thin transition region where the
temperature ramps up to $1.75\times 10^8~\mathrm{K}$ and the
composition changes to 99\% \isot{He}{4} and 1\% \isot{N}{14}.    This He envelope is then integrated isentropically.  The
density of the transition from the underlying CO white dwarf to the He
envelope was selected to yield an envelope mass of $0.05~M_\odot$.

We place a small temperature perturbation in the He layer using the
same form as in \citet{castro_simple_sdc}:
\begin{equation}
  T = T_0 \left [ 1 + X(\isotm{He}{4}) f (1 + \tanh(2 - r_1)) \right ]
\end{equation}
where
\begin{equation}
  r_1 = \left [ x^2 + (y - r_0)^2 \right ]^{1/2} / \lambda
\end{equation}
and
\begin{equation}
  r_0 = r_\mathrm{pert} + r_\mathrm{base}
\end{equation}
Here, $r_\mathrm{base}$ is the radius at which the helium layer begins
and $r_\mathrm{pert}$ is the distance above the base to put the
perturbation.  We choose $r_\mathrm{pert} = 100~\mathrm{km}$.  The
temperature is perturbed above the initial model value, denoted as
$T_0$ here.  The amplitude of the perturbation is $f = 3$ and the
scale of the perturbation is $\lambda = 12.5~\mathrm{km}$, and
$r_\mathrm{pert} = 100~\mathrm{km}$.  This is a very small
perturbation, but enough to seed the initial He detonation in the
envelope.

The simulations are all run on the OLCF Frontier machine, using 4
nodes / 32 AMD GPUs.  The data is moved to the GPUs at the start of
the simulation and all computation is done there.  Our GPU offloading
strategy takes advantage of the \amrex\ C++ lambda-capturing
functionality to be performance portable.

@@

@@

Figure~\ref{fig:lap_rho_sequence} shows a Schlieren-style plot, which
highlights density gradients.  It is constructed by plotting
$\log_{10}(|\rho^{-1}\nabla^2\rho|)$.  This clearly shows the
compression wave launched by the He detonation propagating in toward
the center of the white dwarf.  The middle row shows this compression
wave converging slightly off-center which then leads to the ignition
of the C detonation.  This also highlights how the speed of the He
detonation is important in determining how off-center the C detonation
ignition will be.  Our choice of network was motivated by ensuring
that we capture the energy release from He burning accurately, in
particular the inclusion of the
$\isotm{C}{12}(p,\gamma)\isotm{N}{13}(\alpha,p)\isotm{O}{16}$
sequence.

@@

>>>>

We considered an extreme version of the double detonation scenario in
\citet{castro_simple_sdc}---a very large perturbation was applied to
drive a detonation directly into the underlying carbon-oxygen white
dwarf.  We showed that the Strang split integration method had
difficulty with the integration unless we used tighter tolerances,
while the simplified-SDC method was well-behaved.  Here we look at a
more realistic setup for a double detonation, with a small
perturbation that drives a He detonation around the star and a
compression wave that results in the ignition of a C detonation near
the center of the star.  We focus on the nucleosynthesis and the
efficiency of the method.

@@

We use the freely-available \castro\ simulation code for all the
simulations shown here.  Our domain uses a 2D axisymmetric geomety
with a size of $5.12\times 10^9~\mathrm{cm}$ by $1.024\times
10^{10}~\mathrm{cm}$.  This is much larger than the initial size of
the star, giving it plenty of room to expand (we note that the figures
shown in the results section are all zoomed in on the star).  The
coarse grid is $640\times 1280$ zones and we use 2 levels of
refinement, each a jump of $2\times$, giving a maximum resolution of
$20~\mathrm{km}$.  \castro\ uses the \amrex\ library
\citep{amrex_joss} to manage the grid hierarchy.  The refinement
strategy is picked to refine on the star and any regions where the
temperature is greater than $10^8~\mathrm{K}$.  Finally, \castro\ uses
subcycling, so finer grids are advanced at a smaller timestep than the
coarse grids.

As noted above, we use the unsplit PPM algorithm for advection, with
the Riemann solver from \citet{colellaglaz:1985} and the general
stellar equation of state of \citet{timmes_swesty:2000}.  In
all simulations, the CFL constraint is the
only timestep constraint used.  Self-gravity is done using a full
Poisson solve using geometric multigrid, with Dirichlet boundary
conditions on the domain boundary computed via a multipole expansion
with a maximum order of 6.  For the low density regions outside of the
star, we use a sponge term on the momentum equation to prevent the
very low density material that is not in hydrostatic equilibrium from
raining down on the star.

All simulations use the same reaction network: the {\tt subch\_simple}
as described in \cite{zhi2023}.  This includes 22 nuclei and 94 rates
from ReacLib \citep{reaclib} and is produced with
\pynucastro\ \citep{pynucastro2}.  \MarginPar{show I show the
  network?} Importantly, we include
$\isotm{N}{14}(\alpha,\gamma)\isotm{F}{18}(\alpha,p)\isotm{Ne}{21}$
which creates protons to allow for
$\isotm{C}{12}(p,\gamma)\isotm{N}{13}(\alpha,p)\isotm{O}{16}$.  This
sequence can be faster than
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$, as pointed out by
\citet{shenbildsten} and is important for getting the detonation speed
correctly.  To reduce the size of the network, we have
\pynucastro\ combine some $(\alpha,p)(p,\gamma)$ and $(\alpha,\gamma)$
rates into an effective $(\alpha,\gamma)$ rate.  Screening is provided
following the procedure in \citet{wallace:1982}, combining the
screening functions of \citet{graboske:1973,alastuey:1978,itoh:1979}.

The reaction system is integrated using a version of the VODE ODE
integrator \citep{vode} ported to C++.  Our modifications to this
integrator are described in \citet{castro_simple_sdc}.  We use an
analytic approximation to the Jacobian (the main approximation is that
the species derivatives of the screening function are not included)
and found that we get best results when we disable the caching of the
Jacobian in the integrator.

@@

We will run a suite of simulations all using the same initial model,
constructed following the methodology in \citet{subchandra}.  We use
an isothermal core of C/O of $1.1~M_\odot$ with $T = 10^7~\mathrm{K}$
and a thin transition region where the temperature ramps up to
$1.75\times 10^8~\mathrm{K}$ and the composition changes to 99\%
\isot{He}{4} and 1\% \isot{N}{14}.  This He envelope is then
integrated isentropically.  The density of the transition from the
underlying CO white dwarf to the He envelope was selected to yield an
envelope mass of $0.05~M_\odot$.  The code for generating the model is
freely available at
\url{https://github.com/amrex-astro/initial_models}.  \MarginPar{should
  we do a zenodo release of this?}

We place a small temperature perturbation in the He layer using the
same form as in \citet{castro_simple_sdc}:
\begin{equation}
  T = T_0 \left \{ 1 + X(\isotm{He}{4}) f \left [1 + \tanh(2 - \sigma) \right ] \right \}
\end{equation}
where
\begin{equation}
  \sigma = \left [ x^2 + (y - R_0)^2 \right ]^{1/2} / \lambda
\end{equation}
and
\begin{equation}
  R_0 = R_\mathrm{pert} + R_\mathrm{base}
\end{equation}
where, $R_\mathrm{base}$ is the radius at which the helium layer
begins and $R_\mathrm{pert}$ is the distance above the base to put the
perturbation.  We choose $R_\mathrm{pert} = 100~\mathrm{km}$.  The
temperature is perturbed above the initial model value, denoted as
$T_0$ here.  The amplitude of the perturbation is $f = 3$ and the
scale of the perturbation is $\lambda = 12.5~\mathrm{km}$.  This is a
very small perturbation, but enough to seed the initial He detonation
in the envelope.  \MarginPar{give the peak temperature}

Our base simulation uses the simplified-SDC integration, a CFL number
of $0.2$, and sets $\epsilon_\mathrm{rel} = \epsilon_\mathrm{abs} =
10^{-5}$ for both the species and energy.  Overall, the simulation
requires 21,784 coarse grid timesteps to evolve to $1.0~\mathrm{s}$ of
simulation time.

The simulations are all run on the OLCF Frontier machine, using 4
nodes / 32 AMD GPUs.  The data is moved to the GPUs at the start of
the simulation and all computation is done there.  Our GPU offloading
strategy \citep{castro_gpu} takes advantage of the \amrex\ C++
lambda-capturing functionality to be performance portable.
All of the simulation code is on github at
\url{https://github.com/amrex-astro/} and the inputs files, global
diagnostics used to make the line plots shown below, and metadata
describing the git hashes, compiler environment, and runtime
parameters for all of the simulaitons is available at \MarginPar{add
  Zenodo ref}.

@@

Figure~\ref{fig:temp_sequence} shows the evolution of the temperature field for the base simulation.  The initial perturbation is placed at the north pole at $t=0$ and leads to a detonation in the He layer that begins to wrap around the star.  By $0.1~\mathrm{s}$ the detonation is almost halfway around the star and by $0.2~\mathrm{s}$ is almost at the south pole.  Figure~\ref{fig:lapl_rho_sequence} shows shows a Schlieren-style plot, which
highlights density gradients.  It is constructed by plotting
$\log_{10}(|\rho^{-1}\nabla^2\rho|)$.  This clearly shows a
compression wave launched by the early evolution of the He detonation propagating inward toward
the center of the white dwarf.

@@

Between $t = 0.3$ and $t = 0.7~\mathrm{s}$ the burning is confined to
the outer layer, but the compression wave continues to advance inward,
converging slightly off-center at around $t = 0.7~\mathrm{s}$.  This
highlights how the speed of the He detonation is important in
determining how off-center the C detonation ignition will be.  Our
choice of network was motivated by ensuring that we capture the energy
release from He burning accurately, in particular the inclusion of the
$\isotm{C}{12}(p,\gamma)\isotm{N}{13}(\alpha,p)\isotm{O}{16}$
sequence.

In the bottom row of Figure~\ref{fig:temp_sequence}, we see
the second detonation ignite near the center of the star and
begin to propagate outward.  By $1.0~\mathrm{s}$ it has evolved
past the initial radius of the star and is into the extended
ash layer left behind by the previous He detonation.

@@

We now look at how Strang integration to this base simulation.
